---
title: "Log Gaussian Cox Process (LGCP) analysis"
author: "Alexandra Strang"
date: "`r Sys.Date()`"
output: pdf_document
toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "D:/Points_test")
```

## 1.0 Document set up

This document outlines the analysis part of "Intra-seasonal variation and factors influencing satellite-derived population indices of Ad\'{e}lie penguins (*Pygoscelis adeliae*) in the Ross Sea", the first data chapter of Alexandra Strang's doctoral thesis. This document focuses on the second question of this chapter: What factors influence the ability to use guano area extracted from VHR imagery to estimate the number of breeding pairs (i.e., nest density, terrain slope, terrain roughness, etc.)? We hypothesised that the inclusion of fine-scale terrain characteristics (i.e., terrain slope and roughness) and nesting densities will improve predictions of the number of breeding pairs from VHR-derived guano area. We aimed to develop a point process model (a Log Gaussian Cox Process (LGCP) model) to relate the location of breeding pairs from drone imagery surveys to guano areas derived from VHR imagery and fine-scale terrain characteristics. Then, theoretically, we could apply our model to a new site where drone surveys have not been done but terrain variables and guano areas have been estimated and predict the number of breeding pairs. 

```{r packages, results='hide'}

# load packages
library(sf)
library(fmesher)
library(ggplot2)
library(INLA)
library(ggpubr)
library(inlabru) # for lgcp()
library(dplyr)
library(purrr)
library(tidyr)
library(terra) # for rasters

# sessionInfo() - important to note r version and versions of inlabru, INLA, fmesher
# R version 4.5.0 (2025-04-11 ucrt)

```

## 2.0 Import data

### 2.1 Penguin nest locations (points)

We use the observed number of breeding pairs obtained by drone survey for Cape Crozier Ad\'{e}lie penguin colony in 2020 (2020-11-29) as our point locations. 

```{r penguin-points}

# read in points from xy csv
# Cape Crozier 2020
Crozier_xy <- read.csv("Crozier_Points_2020_3031.csv")

# convert to sf object and check CRS
sf_Crozier <- st_as_sf(Crozier_xy, coords = c("x", "y"), crs = 3031)
st_crs(sf_Crozier)

# EPSG 3031 is in metres but when plotting geom_sf will convert to degrees

# plot points
point_plot_Crozier <- ggplot() + 
  geom_sf(data = sf_Crozier, color = "purple", size = 1.7, alpha = 0.5) + 
  labs( 
    x = "Easting", 
    y = "Northing", 
    ) + 
  theme_minimal()
# geom_sf converts to degrees

point_plot_Crozier

# eventually use Cape Crozier 2019 GA and points
# for model training as it has better precison and
# foercasts rather than hindcasts

```

### 2.2 VHR-derived guano area

Visualise the area of guano derived from VHR imagery at Cape Crozier for 2020. 

```{r guano-area}

# import guano area shapefile
sf_Crozier_guano <- st_read("Point_process_GA_boundaries/Crozier_2020_1_3031_guano.shp")
# ensure shapefile has right crs code
sf_Crozier_guano <- st_transform(sf_Crozier_guano, crs = st_crs(sf_Crozier))

# plot guano area
GA_plot_Crozier <- ggplot() + 
  geom_sf(data = sf_Crozier_guano, fill = NA, color = "orange") + 
  labs( 
    x = "Easting", 
    y = "Northing", 
    ) + 
  theme_minimal()
# geom_sf converts to degrees

GA_plot_Crozier

# show points and guano area together
Together <- plot(ggarrange(point_plot_Crozier, 
                           GA_plot_Crozier,
                        ncol = 2, nrow = 1, labels=c("a","b")))
#annotate_figure(Together, left = "Northing", bottom = "Easting")

```

### 2.3 Model domain (coastline boundary)

Use the boundary of the coastline at Cape Crozier and the bounds of the VHR image to create the boundary (domain) to be used to create INLA (Integrated Nested Laplace Approximation) mesh for the LGCP model. This domain specifies the area to focus the model calculations within (excludes the area beyond the coastline as we know there are no nests present there).

```{r coastline-boundary}

# import Crozier coastline boundary
sf_Crozier_boundary <- st_read("Crozier_boundary.shp")
# transform shapefile so it has the right crs code
sf_Crozier_boundary <- st_transform(sf_Crozier_boundary, crs = st_crs(sf_Crozier))
st_crs(sf_Crozier_boundary)

# plot coastline boundary
Boundary_plot_Crozier <- ggplot() + 
  geom_sf(data = sf_Crozier_boundary, fill = NA, color = "blue", linetype = "dashed") +
  labs( 
    x = "Easting", 
    y = "Northing", 
    ) + 
  theme_minimal()
# geom_sf converts to degrees

Boundary_plot_Crozier

```

### 2.4 Model samplers area (drone image boundary)

Use the boundary of the drone image of Cape Crozier in 2020 (2020-11-29) to create the samplers area which is the area that has been 100% sampled for points (penguin nests).

```{r UAV-boundary}

# import Crozier UAV bounds shapefile
sf_Crozier_UAV_area <- st_read("Crozier_20201129_UAV/Crozier_20201129_3031_UAV_area.shp")
# ensure shapefile has right crs code
sf_Crozier_UAV_area <- st_transform(sf_Crozier_UAV_area, crs = st_crs(sf_Crozier))
st_crs(sf_Crozier_UAV_area)

# plot UAV boundary
UAV_plot_Crozier <- ggplot() + 
  geom_sf(data = sf_Crozier_UAV_area, fill = NA, color = "red") + 
  labs( 
    x = "Easting", 
    y = "Northing", 
    ) + 
  theme_minimal()
# geom_sf converts to degrees

UAV_plot_Crozier

```

## 3.0 Creating the mesh

```{r mesh}

# mesh parameters
# this is 1/15 study size using x range
# x range is longer than y here
Crozier_max.edge <- diff(range(st_coordinates(sf_Crozier)[,1]))/(3*5)
print(Crozier_max.edge)
# ~150 metres

# expand outer layer out by 1/5
Crozier_bound.outer = diff(range(st_coordinates(sf_Crozier)[,1]))/5
print(Crozier_bound.outer)
# ~500 metres

# create Crozier mesh
Crozier_mesh <- fm_mesh_2d(boundary = sf_Crozier_boundary, # use coastline as boundary
                            max.edge = c(1,5)*Crozier_max.edge, # inner and outer max edge where outer layer has triangle density lower than inner
                            offset = c(Crozier_max.edge, Crozier_bound.outer),
                            cutoff = Crozier_max.edge/10,
                            crs = st_crs(sf_Crozier))

# mesh is exported as shapefile to crop terrain covariates to mesh boundary

# plot boundary mesh
mesh_plot_Crozier <- ggplot() + 
  geom_fm(data = Crozier_mesh) + 
  geom_sf(data = sf_Crozier_boundary, fill = NA, color = "blue", linetype = "dashed") + 
  labs( 
    x = "Easting", 
    y = "Northing", 
    ) + 
  theme_minimal()
# geom_sf converts to degrees

mesh_plot_Crozier

```

## 4.0 LGCP model fitting

```{r LGCP_null}

# create the spatial random field
# fit an LGCP model to the locations of the penguin nests

# Penalized Complexity (PC) priors
# the probability of the range exceeding that to be 0.5
# the prior for the variance explained by the spatial effect
# is set that the probability that the SD exceeds 1 is 0.5

# define the SPDE priors (matern)
matern <- inla.spde2.pcmatern(mesh = Crozier_mesh,
                              prior.range = c(250, 0.5), # distance decay in metres
                              prior.sigma = c(1, 0.5)) # amount of spatial variation

# formula specification of model components
# specify a model where for 2D models geometry is on the left of ~
# and an SPDE + Intercept(1) on the right
# define the domain of the LGCP and model components
# (spatial SPDE effect and Intercept)
null_cmp <- geometry ~
  Intercept(1) + # fixed effect (eventually add here the covariates of slope etc. using rasters)
  mySmooth(geometry, model = matern) # random effect

# formula (geometry is response = point locations)
# mysmooth for spatial autocorrelation

# use lgcp() with 2D model components, the sf points and the sf boundary
null_model <- lgcp(null_cmp, # formula
                    data = sf_Crozier, # locations
                    samplers = sf_Crozier_UAV_area, # sample area of UAV bounds
                    domain = list(geometry = Crozier_mesh), # mesh
                    options = list(
                      control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
                    )
)
# takes less than ~2 mins to run

summary(null_model)
# mean should be density per m 2 (once unlogged)

# predicting intensity 
# predict the spatial intensity surface
lambda <- predict(
  null_model, 
  fm_pixels(Crozier_mesh, format = "sf", mask = sf_Crozier_UAV_area), # use samplers 
  ~ exp(mySmooth + Intercept) # exp to unlog, mySmooth is field
  )
# sum of mean just gives density

# need to sum over prediction grid
# make prediction grid as sf points
grid_pts <- fm_pixels(
  Crozier_mesh,
  format = "sf",
  mask = sf_Crozier_UAV_area
)

# predict intensity per m2
lambda <- predict(
  null_model,
  grid_pts, # use this instead of fm_pixels
  ~ exp(mySmooth + Intercept)
)

# cell spacing from the point coordinates
coords <- st_coordinates(grid_pts)
dx <- median(diff(sort(unique(coords[,1]))))
dy <- median(diff(sort(unique(coords[,2]))))

cell_area <- dx * dy  # m2 per pixel

# multiply and sum
total_pred <- sum(lambda$mean * cell_area)
total_pred

# plot log intensity
Intensity_plot <- ggplot() +
  geom_fm(data = Crozier_mesh) +
  gg(lambda, geom = "tile")

Intensity_plot
# intensity is highest at 0.5 ish (seems right for density, maybe a little low)

```

## 5.0 Analyse model sensitivity to prior specifications

```{r prior_sensitivity}

# analyse sensitivity of model to prior specifications

# start with the prior range first

# prior ranges to test (100, 250, 500, 750, 1000 m)
range_grid <- tibble(
  r0 = c(100, 250, 500, 750, 1000),  # prior ranges
  pr = 0.5,                       # probability that range < r0
  s0 = 1,                         # baseline sigma anchor
  ps = 0.5                        # prob(sigma > s0)
)

# function for fitting LGCP models for range priors
fit_LGCP <- function(r0, pr, s0, ps) {
  matern <- inla.spde2.pcmatern(
    mesh = Crozier_mesh,
    prior.range = c(r0, pr),
    prior.sigma = c(s0, ps)
  )
  
  cmp <- geometry ~ Intercept(1) + mySmooth(geometry, model = matern)
  
  model <- lgcp(
    cmp,
    data = sf_Crozier,
    samplers = sf_Crozier_UAV_area,
    domain = list(geometry = Crozier_mesh),
    options = list(control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE))
  )
  
  lambda <- predict(model, grid_pts, ~ exp(mySmooth + Intercept))
  total_pred <- sum(lambda$mean * cell_area)
  
  # hyperparameter summaries
  hp <- model$summary.hyperpar
  get_row <- function(pattern) {
    hp[grepl(pattern, rownames(hp), ignore.case = TRUE), , drop = FALSE]
  }
  range_row <- get_row("range")
  stdev_row <- get_row("stdev|std")
  
  tibble(
    r0 = r0, pr = pr, s0 = s0, ps = ps,
    WAIC = model$waic$waic,
    DIC = model$dic$dic,
    post_range_mean = range_row$mean[1],
    post_range_2.5 = range_row$`0.025quant`[1],
    post_range_97.5 = range_row$`0.975quant`[1],
    post_sd_mean = stdev_row$mean[1],
    post_sd_2.5 = stdev_row$`0.025quant`[1],
    post_sd_97.5 = stdev_row$`0.975quant`[1],
    total_pred = total_pred
  )
}

# Combine results
results_range <- pmap_dfr(range_grid, fit_LGCP)
print(results_range)

# plots

# Posterior Range vs prior r0
ggplot(results_range, aes(x = r0, y = post_range_mean)) +
  geom_point(size = 3) +
  geom_line() +
  geom_errorbar(aes(ymin = post_range_2.5, ymax = post_range_97.5), width = 30) +
  labs(x = "Prior range anchor r0 (m)",
       y = "Posterior mean range (95% CI)",
       title = "Posterior spatial range vs prior range") +
  theme_minimal()

# Posterior SD vs prior r0
ggplot(results_range, aes(x = r0, y = post_sd_mean)) +
  geom_point(size = 3) +
  geom_line() +
  geom_errorbar(aes(ymin = post_sd_2.5, ymax = post_sd_97.5), width = 30) +
  labs(x = "Prior range anchor r0 (m)",
       y = "Posterior SD (95% CI)",
       title = "Posterior SD vs prior range") +
  theme_minimal()

# WAIC vs prior r0
ggplot(results_range, aes(x = r0, y = WAIC)) +
  geom_point(size = 3) +
  geom_line() +
  labs(x = "Prior range anchor r0 (m)",
       y = "WAIC",
       title = "Model WAIC vs prior range") +
  theme_minimal()

# DIC vs prior r0
ggplot(results_range, aes(x = r0, y = DIC)) +
  geom_point(size = 3) +
  geom_line() +
  labs(x = "Prior range anchor r0 (m)",
       y = "DIC",
       title = "Model DIC vs prior range") +
  theme_minimal()

# Total predicted count vs prior r0
ggplot(results_range, aes(x = r0, y = total_pred)) +
  geom_point(size = 3) +
  geom_line() +
  labs(x = "Prior range anchor r0 (m)",
       y = "Total predicted count",
       title = "Predicted total vs prior range") +
  theme_minimal()

```

## 6.0 Covariates

### 6.1 Import rasters

```{r import_rasters}

# Build LGCP model with covarites

# Incorporate guano area and terrain variables as rasters


```

### 6.2 Standardise covarites

```{r standardise_covariates}

# scale
# (mean of 0 sd of 1)


```

### 6.3 Collinearity

```{r check_collinearity}


```

### 6.4 Fit LGCP

```{r LGCP_covariates}




```

